Programming Language Foundations in Agda

### 2019-07-21 papers/scp/PLFA.tex
switch to extrinsic/intrinsic terminology

% WRITE TO JEREMY RE: DIFFERENCE IN STYLE OF CHAPTER TITLES
% NEED TO CITE SBMF PAPER AND DESCRIBE DIFFERENCES
% STRIP OUT EXAM RESULTS
% CHANGE INHERENT --> INTRINSIC, CITE REYNOLDS

@@

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using raw terms with a separate typing relation is far less
  perspicuous than using inherently-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

Section~5 claims that raw terms should be avoided in favour of
inherently-typed terms.  PLFA develops lambda calculus with both raw
and inherently-typed terms, permitting a comparison.  It turns out the
former is less powerful---it supports substitution only for closed
terms---but significantly longer---about 1.6 times as many lines of code,
roughly the golden ratio.

@@

\paragraph{Lambda: Introduction to lambda calculus}
Introduces lambda calculus, using a representation with named
variables and a separate typing relation. The language used is PCF
\citep{Plotkin-1977}, with variables, lambda abstraction, application,
zero, successor, case over naturals, and fixpoint. Reduction is
call-by-value and restricted to closed terms.

@@

\paragraph{DeBruijn: Inherently typed de Bruijn representation}
Introduces de Bruijn indices and the inherently-typed representation.
Emphasis is put on the structural similarity between a term and its
corresponding type derivation; in particular, de Bruijn indices
correspond to the judgment that a variable is well-typed under a given
environment.

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types; and explains lists and let bindings.
Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The inherently typed representation is used.

@@

\paragraph{Inference: Bidirectional type inference}
Introduces bidirectional type inference, and applies it to convert from
a representation with named variables and a separate typing relation
to a representation de Bruijn indices with inherent types. Bidirectional
type inference is shown to be both sound and complete.

\paragraph{Untyped: Untyped calculus with full normalisation}
As a variation on earlier themes, discusses an untyped (but inherently
scoped) lambda calculus.  Reduction is call-by-name over open terms,
with full normalisation (including reduction under lambda terms).  Emphasis
is put on the correspondence between the structure of a term and
evidence that it is in normal form.

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an inherently-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.  The new part on
Denotational Semantics also covers material not treated by SF.

@@

There is more material that would be desirable to include in PLFA which was not
due to limits of time, including mutable references, logical relations, System F, and
pure type systems. We'd especially like to include pure type systems as they
provide the readers with a formal model close to the dependent types used in the
book.  Our attempts so far to formalise pure type systems have proved
challenging, to say the least.

@@

% TODO: rewrite use of 'Philip' to reflect all authors?
Philip was therefore surprised to realise that any constructive proof of
progress and preservation \emph{automatically} gives rise to such a
prototype.  The input is a term together with evidence the term is
well-typed.  (In the inherently-typed case, these are the same thing.)
Progress determines whether we are done, or should take another step;
preservation provides evidence that the new term is well-typed, so we
may iterate. In a language with guaranteed termination, we cannot
iterate forever, but there are a number of well-known techniques to
address that issue; see, e.g., \citet{Bove-and-Capretta-2001},
\citet{Capretta-2005}, or \citet{McBride-2015}.
We use the simplest, similar to McBride's \emph{petrol-driven} (or
\emph{step-indexed}) semantics: provide a maximum number of steps to
execute; if that number proves insufficient, the evaluator returns the
term it reached, and one can resume execution by providing a new
number.

@@

\section{Inherent typing is golden}

\begin{figure}[p]
  \includegraphics[width=\textwidth]{figures/raw.png}
  \caption{Raw approach in PLFA}
  \label{fig:raw}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=\textwidth]{figures/inherent.png}
  \caption{Inherent approach in PLFA}
  \label{fig:inherent}
\end{figure}

% TODO: consider rewriting this?
The second part of PLFA first discusses two different approaches to
modeling simply-typed lambda calculus.  It first presents raw
terms with named variables and a separate typing relation and
then shifts to inherently-typed terms with de Bruijn indices.
Before writing the text, Philip had thought the two approaches
complementary, with no clear winner.  Now he is convinced that the
inherently-typed approach is superior.

Figure~\ref{fig:raw} presents the raw approach.
It first defines $\texttt{Id}$, $\texttt{Term}$,
$\texttt{Type}$, and $\texttt{Context}$, the abstract syntax
of identifiers, raw terms, types, and contexts.
It then defines two judgments,
$\GG\:{\ni}\:\ex\:{\typecolon}\:\AY$ and
$\GG\:{\vdash}\:\EM\:{\typecolon}\:\AY$,
which hold when under context $\GG$ the variable $\ex$
and the term $\EM$ have type $\AY$, respectively.

Figure~\ref{fig:inherent} presents the inherent approach.
It first defines $\texttt{Type}$ and $\texttt{Context}$, the abstract syntax
of types and contexts, of which the first is as before and the second is
as before with identifiers dropped.  In place of the two judgments,
the types of variables and terms are indexed by a context and a type,
so that $\GG\:{\ni}\:\AY$ and $\GG\:{\vdash}\:\AY$ denote
variables and terms, respectively, that under context $\GG$ have type $\AY$.
The indexed types closely resemble the previous judgments:
we now represent a variable or a term by the proof that it is well typed.
In particular, the proof that a variable is well typed in the raw approach
corresponds to a de Bruijn index in the inherent approach.

The raw approach requires more lines of code than the inherent approach.  The
separate definition of raw terms is not needed in the inherent approach; and one
judgment in the raw approach needs to check that $\ex \not\equiv \why$, while
the corresponding judgment in the inherent approach does not.  The difference
becomes more pronounced when including the code for substitution, reductions,
and proofs of progress and preservation.  In particular, where the raw approach
requires one first define substitution and reduction and then prove they
preserve types, the inherent approach establishes substitution at the same time
it defines substitution and reduction.

Stripping out examples and any proofs that appear in one but not the other (but
could have appeared in both), the full development in PLFA for the raw approach
takes 451 lines (216 lines of definitions and 235 lines for the proofs) and the
development for the inherent approach takes 275 lines (with definitions and
proofs interleaved).  We have 451 / 235 = 1.64, close to the golden ratio.

The inherent approach also has more expressive power.  The raw
approach is restricted to substitution of one variable by a closed
term, while the inherent approach supports simultaneous substitution
of all variables by open terms, using a pleasing formulation due to
\citet{McBride-2005}, inspired by \citet{Goguen-and-McKinna-1997} and
\citet{Altenkirch-and-Reus-1999} and described in
\citet{Allais-et-al-2017}. In fact, we did manage to write a variant of
the raw approach with simultaneous open substitution along the lines
of McBride, but the result was too complex for use in an introductory
text, requiring 695 lines of code---more than the total for the other
two approaches combined.
% TODO: is this still available somewhere in extra/?

The text develops both approaches because the raw approach is more
familiar, and because placing the inherent approach first would
lead to a steep learning curve.  By presenting the more long-winded
but less powerful approach first, students can see for themselves the
advantages of de Bruijn indices and inherent types.

There are actually four possible designs, as the choice of named
variables vs de Bruijn indices, and the choice of raw vs
inherently-typed terms may be made independently.  There are synergies
between the two.  Manipulation of de Bruijn indices can be notoriously
error-prone without inherent-typing to give assurance of correctness.
In inherent typing with named variables, simultaneous substitution by
open terms remains difficult.

The benefits of the inherent approach 
are well known to some. The technique was introduced by
\citet{Altenkirch-and-Reus-1999}, and widely used elsewhere,
notably by \citet{Chapman-2009} and \citet{Allais-et-al-2017}.
Philip is grateful to David Darais for bringing it to his attention.

@@

Students must answer question~1 on the exam, and one of questions~2
or~3.  Question~1 gives predicates over a data structure, such lists or
trees, to be formalised in Agda, and a theorem relating the
predicates, to be formalised and proved in Agda.  Question~2 gives the
students the inherent formulation of lambda calculus from chapter
DeBruijn, which they must extend with a described language feature.
Question~3 give the students the the bidirectional type inferencer
from chapter Inference, which they must extend with a described
language feature.

@@

% Table~\ref{tab:exams} shows performance on the mock exam and the final
% exam for the courses run at Edinburgh and PUC-Rio.  While scores vary
% widely on the mock, all students achieve perfection on the
For the courses run at Edinburgh and PUC-Rio, the scores vary
widely on the mock: minimum <20%, maximum 100%, mean
77.8, standard deviation 27.6. But all students achieve perfection on the
exam.  (The one exception was a PUC-Rio student who did not attend
classes or sit the mock.)  Similar results were achieved at Edinburgh
over the previous five years, using SF as the course textbook and Coq
as the proof assistant.  We consider these results a tribute to the
students' ability to study and learn.

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing how
much more compact it is to avoid raw terms, thanks to David Darais.  For
inspiring our work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.

>>>>

@@

  What did we learn from writing PLFA? First, that it is possible. One
  might expect that without proof tactics that the proofs become too
  long, but in fact proofs in PLFA are about the same length as those
  in SF. Proofs in Coq require an interactive environment to be
  understood, while proofs in Agda can be read on the page.  Second,
  that constructive proofs of preservation and progress give immediate
  rise to a prototype evaluator. This fact is obvious in retrospect
  but it is not exploited in SF (which instead provides a separate
  normalise tactic) nor can I find it in the literature.  Third, that
  using extrinsicly-typed terms is far less
  perspicuous than using intrinsicly-typed terms. SF uses the former
  presentation, while PLFA presents both; the former uses about 1.6 as
  many lines of Agda code as the latter, roughly the golden ratio.

@@

Section~5 claims that extrinsicly-typed terms should be avoided in favour of
intrisicly-typed terms.  PLFA develops lambda calculus with both,
permitting a comparison.  It turns out the
former is less powerful---it supports substitution only for closed
terms---but significantly longer---about 1.6 times as many lines of code,
roughly the golden ratio.

@@

\paragraph{Lambda: Introduction to lambda calculus}
Introduces lambda calculus, using a representation with named
variables and extrinsicly typed. The language used is PCF
\citep{Plotkin-1977}, with variables, lambda abstraction, application,
zero, successor, case over naturals, and fixpoint. Reduction is
call-by-value and restricted to closed terms.

@@

\paragraph{DeBruijn: Intrinsicly-typed de Bruijn representation}
Introduces de Bruijn indices and the intrinsicly-typed representation.
Emphasis is put on the structural similarity between a term and its
corresponding type derivation; in particular, de Bruijn indices
correspond to the judgment that a variable is well-typed under a given
environment.

\paragraph{More: More constructs of simply-typed lambda calculus}
Introduces product, sum, unit, and empty types; and explains lists and let bindings.
Typing and reduction rules are given informally; a few
are then give formally, and the rest are left as exercises for the reader.
The intrinsicly-typed representation is used.

@@

\paragraph{Inference: Bidirectional type inference}
Introduces bidirectional type inference, and applies it to convert from
a representation with named variables and extrinsicly typed
to a representation with de Bruijn indices and intrinsicly typed. Bidirectional
type inference is shown to be both sound and complete.

\paragraph{Untyped: Untyped calculus with full normalisation}
As a variation on earlier themes, discusses an untyped (but intrinsicly
scoped) lambda calculus.  Reduction is call-by-name over open terms,
with full normalisation (including reduction under lambda terms).  Emphasis
is put on the correspondence between the structure of a term and
evidence that it is in normal form.

@@

PLFA and SF differ in several particulars.  PLFA begins with a computationally
complete language, PCF, while SF begins with a minimal language, simply-typed
lambda calculus with booleans.  PLFA does not include type annotations in terms,
and uses bidirectional type inference, while SF has terms with unique types and
uses type checking.  SF also covers a simple imperative language with Hoare
logic, and for lambda calculus covers subtyping, record types, mutable
references, and normalisation---none of which are treated by PLFA.  PLFA covers
an intrinsicly-typed de Bruijn representation, bidirectional type inference,
bisimulation, and an untyped call-by-name language with full
normalisation---none of which are treated by SF.  The new part on
Denotational Semantics also covers material not treated by SF.

@@

There is more material that would be desirable to include in PLFA which was not
due to limits of time, including mutable references, logical relations, System~F, and
pure type systems. We'd especially like to include pure type systems as they
provide the readers with a formal model close to the dependent types used in the
book.  Our attempts so far to formalise pure type systems have proved
challenging, to say the least.

@@

% TODO: rewrite use of 'Philip' to reflect all authors?
Philip was therefore surprised to realise that any constructive proof of
progress and preservation \emph{automatically} gives rise to such a
prototype.  The input is a term together with evidence the term is
well-typed.  (In the intrinsicly-typed case, these are the same thing.)
Progress determines whether we are done, or should take another step;
preservation provides evidence that the new term is well-typed, so we
may iterate. In a language with guaranteed termination such as Agda, we cannot
iterate forever, but there are a number of well-known techniques to
address that issue; see, e.g., \citet{Bove-and-Capretta-2001},
\citet{Capretta-2005}, or \citet{McBride-2015}.
We use the simplest, similar to McBride's \emph{petrol-driven} (or
\emph{step-indexed}) semantics: provide a maximum number of steps to
execute; if that number proves insufficient, the evaluator returns the
term it reached, and one can resume execution by providing a new
number.

@@

\section{Intrinsic typing is golden}

\begin{figure}[p]
  \includegraphics[width=\textwidth]{figures/raw.png}
  \caption{Extrinsic approach in PLFA}
  \label{fig:extrinsic}
\end{figure}

\begin{figure}[t]
  \includegraphics[width=\textwidth]{figures/inherent.png}
  \caption{Intrinsic approach in PLFA}
  \label{fig:intrinsic}
\end{figure}

% TODO: consider rewriting this?
The second part of PLFA first discusses two different approaches to
modeling simply-typed lambda calculus.  It first presents
terms with named variables and and extrinsic typing relation and
then shifts to terms with de Bruijn indices that are intrinsicly
typed.  The names \emph{extrinsic} and \emph{intrinsic} for these
two approaches are taken from \citet{Reynolds-2003}.
Before writing the text, Philip had thought the two approaches
complementary, with no clear winner.  Now he is convinced that the
intrinsicly-typed approach is superior.

Figure~\ref{fig:extrinsic} presents the extrinsic approach.
It first defines $\texttt{Id}$, $\texttt{Term}$,
$\texttt{Type}$, and $\texttt{Context}$, the abstract syntax
of identifiers, raw terms, types, and contexts.
It then defines two judgments,
$\GG\:{\ni}\:\ex\:{\typecolon}\:\AY$ and
$\GG\:{\vdash}\:\EM\:{\typecolon}\:\AY$,
which hold when under context $\GG$ the variable $\ex$
and the term $\EM$ have type $\AY$, respectively.

Figure~\ref{fig:intrinsic} presents the intrinsic approach.
It first defines $\texttt{Type}$ and $\texttt{Context}$, the abstract syntax
of types and contexts, of which the first is as before and the second is
as before with identifiers dropped.  In place of the two judgments,
the types of variables and terms are indexed by a context and a type,
so that $\GG\:{\ni}\:\AY$ and $\GG\:{\vdash}\:\AY$ denote
variables and terms, respectively, that under context $\GG$ have type $\AY$.
The indexed types closely resemble the previous judgments:
we now represent a variable or a term by the proof that it is well typed.
In particular, the proof that a variable is well typed in the extrinsic approach
corresponds to a de Bruijn index in the intrinsic approach.

The extrinsic approach requires more lines of code than the intrinsic approach.  The
separate definition of raw terms is not needed in the intrinsic approach; and one
judgment in the extrinsic approach needs to check that $\ex \not\equiv \why$, while
the corresponding judgment in the intrinsic approach does not.  The difference
becomes more pronounced when including the code for substitution, reductions,
and proofs of progress and preservation.  In particular, where the extrinsic approach
requires one first define substitution and reduction and then prove they
preserve types, the intrinsic approach establishes substitution at the same time
it defines substitution and reduction.

Stripping out examples and any proofs that appear in one but not the other (but
could have appeared in both), the full development in PLFA for the extrinsic approach
takes 451 lines (216 lines of definitions and 235 lines for the proofs) and the
development for the intrinsic approach takes 275 lines (with definitions and
proofs interleaved).  We have 451 / 235 = 1.64, close to the golden ratio.

The intrinsic approach also has more expressive power.  The extrinsic
approach is restricted to substitution of one variable by a closed
term, while the intrinsic approach supports simultaneous substitution
of all variables by open terms, using a pleasing formulation due to
\citet{McBride-2005}, inspired by \citet{Goguen-and-McKinna-1997} and
\citet{Altenkirch-and-Reus-1999} and described in
\citet{Allais-et-al-2017}. In fact, we did manage to write a variant of
the extrinsic approach with simultaneous open substitution along the lines
of McBride, but the result was too complex for use in an introductory
text, requiring 695 lines of code---more than the total for the other
two approaches combined.
% TODO: is this still available somewhere in extra/?

The text develops both approaches because the extrinsic approach is more
familiar, and because placing the intrinsic approach first would
lead to a steep learning curve.  By presenting the more long-winded
but less powerful approach first, students can see for themselves the
advantages of de Bruijn indices and intrinsic types.

There are actually four possible designs, as the choice of named
variables vs de Bruijn indices, and the choice of extrinsic vs
intrinsic typing may be made independently.  There are synergies
between the two.  Manipulation of de Bruijn indices can be notoriously
error-prone without intrinsic types to give assurance of correctness.
For instrinsic typing with named variables, simultaneous substitution by
open terms remains difficult.

The benefits of the intrinsic approach 
are well known to some. The technique was introduced by
\citet{Altenkirch-and-Reus-1999}, and widely used elsewhere,
notably by \citet{Chapman-2009} and \citet{Allais-et-al-2017}.
Philip is grateful to David Darais for bringing it to his attention.

@@

Students must answer question~1 on the exam, and one of questions~2
or~3.  Question~1 gives predicates over a data structure, such lists or
trees, to be formalised in Agda, and a theorem relating the
predicates, to be formalised and proved in Agda.  Question~2 gives the
students the intrinsic formulation of lambda calculus from chapter
DeBruijn, which they must extend with a described language feature.
Question~3 give the students the the bidirectional type inferencer
from chapter Inference, which they must extend with a described
language feature.

@@

% Table~\ref{tab:exams} shows performance on the mock exam and the final
% exam for the courses run at Edinburgh and PUC-Rio.  While scores vary
% widely on the mock, all students achieve perfection on the
For the courses run at Edinburgh and PUC-Rio, the scores vary
widely on the mock: minimum <20\%, maximum 100\%, mean
77.8, standard deviation 27.6. But all students achieve perfection on the
exam.  (The one exception was a PUC-Rio student who did not attend
classes or sit the mock.)  Similar results were achieved at Edinburgh
over the previous five years, using SF as the course textbook and Coq
as the proof assistant.  We consider these results a tribute to the
students' ability to study and learn.

@@

For inventing ideas on which PLFA is based, and for hand-holding, many thanks to
Conor McBride, James McKinna, Ulf Norell, and Andreas Abel.  For showing how
much more compact it is to use intrinsicly-typed terms, thanks to David Darais.  For
inspiring our work by writing SF, thanks to Benjamin Pierce and his coauthors.
For comments on a draft of this paper, an extra thank you to James McKinna, Ulf
Norell, Andreas Abel, and Benjamin Pierce. This research was supported by EPSRC
Programme Grant EP/K034413/1.
